Visualizing Independence Using Extended Association Plots

### 2003-06-10 trunk/vcd/paper/dsc2003.Rnw
more editorial changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@267 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{abstract}
Association plots---a visualization technique for the independence problem in 2-way
contingency tables---are extended in three directions:
\begin{enumerate}
 \item The visualization is enhanced by using colors for the
  importance of the residuals.
 \item Two extensions for the use of multi-way tables are proposed.
 \item The implementation currently available in \textsf{R} is improved using a more modular
  design and allowing a more flexible specification of plotting parameters.
\end{enumerate}
\end{abstract}

@@

Given two categorical variables, or, equivalently, their contingency
table, one is often interested in whether they are 
independent or not. Usually, the $\chi^2$ test statistic---the sum of the
squared Pearson residuals---is used as a measure of association.
Now, whenever the statistic is
significant, one might ask for a more detailed analysis on the basis
of the residuals themselves.
This task can greatly be sustained by the use of graphical
methods such as the association plot, for which we propose some
extensions: the use of colors for the shading of the residuals, two
possible extensions for multi-way tables, and a new, more flexible
implementation using {\sf R}'s alternative graphical system,
\texttt{grid} \cite{vcd:Murrell:2002}. The new functionality is provided by the {\sf R} package
\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
\citep{vcd:Friendly:2000}.

@@

Figure \ref{fig:assoc-classic1} shows the association plot
(produced by the implementation in \texttt{base} {\sf R}) for the
variables `Hair' and `Eye' of the well-known `HairEyeColor' data
set \cite[Hair and Eye colors of 264 male and 328 female statistics
students; see][]{vcd:Snee:1974}. 
The biggest tiles are easily made out: e.g., the tiles for 
brown and blue eyes, given blond hair, seem to represent important residuals. 
But assessing the ``importance'' of the others is far from obvious.

@@

We propose to use a color-shading similar to
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
\cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1984}:
this extension 
uses a color coding of the mosaic tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table.
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

Another example is given in Figure \ref{fig:assoc-classic2}:
this time, we cross-tabulate the variables `Eye' and `Sex'.
For example, the big tiles for blond hair convey the impression of an asymmetry between
male and female students, and hence that hypothesis of independence is
to be rejected. However, the $\chi^2$ test statistic is not significant at
a 5\% level: in this example, its value is 6.221, and the corresponding
$p$ value approximately $0.1013$. Apparently, the relative size of the
tiles is insufficient to assess the ``strength'' of association
between two variables; again, information on the absolute size of the
residuals would be useful.
Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
that \emph{no} residual is bigger than 2 (or smaller than $-2$ for
negative ones, respectively) 
because simply no tile is shaded. In fact, as we can see from
the legend, the overall range of the residuals is smaller than $[-1.5,1.5]$.

@@

To visualize \textit{conditional} independence, we could use conditioning
plots: like in trellis displays, we produce a separate plot
for each level of a conditioning variable (in the case of two
or more conditioning variables, for each level combination). 
This can be illustrated using the famous admissions data of the University of
California at Berkley (UCB)---see \cite{vcd:Bickel+Hammel+O'connell:1975}---which is available in base \textsf{R}.
In this data, the question whether there is gender
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no gender discrimination
in the department strata---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
association plot in Figure \ref{fig:UCB-assoc-signif}.

@@

>>>>

\begin{abstract}
Association plots---a visualization technique for the independence problem in 2-way
contingency tables---are extended in three directions:
\begin{enumerate}
 \item The visualization is enhanced by using colors for the
  importance of the residuals.
 \item Two methods for the use of multi-way tables are proposed.
 \item The implementation currently available in \textsf{R} is improved using a more modular
  design and allowing a more flexible specification of plotting parameters.
\end{enumerate}
\end{abstract}

@@

Given two categorical variables, or, equivalently, their contingency
table, one is often interested in whether they are 
independent or not. Usually, the $\chi^2$ test statistic---the sum of the
squared Pearson residuals---is used as a measure of association.
Now, whenever the statistic is
significant, one might ask for a more detailed analysis on the basis
of the residuals themselves.
This task can greatly be sustained by the use of graphical
methods such as the association plot, for which we propose some
extensions: the use of colors for the shading of the residuals, two
possible methods for multi-way tables, and a new, more flexible
implementation using {\sf R}'s alternative graphical system
\texttt{grid} \citep{vcd:Murrell:2002}. The new functionality is provided 
by the {\sf R} package
\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
\citep{vcd:Friendly:2000}.

@@

Figure \ref{fig:assoc-classic1} shows the association plot
(produced by the implementation in base {\sf R}) for the
variables `Hair' and `Eye' of the well-known `HairEyeColor' data
set \cite[Hair and Eye colors of 264 male and 328 female statistics
students; see][]{vcd:Snee:1974}. 
The biggest tiles are easily made out: e.g., the tiles for 
brown and blue eyes, given blond hair, seem to represent important residuals. 
But assessing the ``importance'' of the others is far from obvious.

@@

We propose to use a color-shading similar to the
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
\cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1984}:
this extension 
uses a color coding of the mosaic tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table.
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

Another example is given in Figure \ref{fig:assoc-classic2}:
this time, we cross-tabulate the variables `Eye' and `Sex'.
For example, the big tiles for blond hair convey the impression of an asymmetry between
male and female students, and hence that hypothesis of independence is
to be rejected. However, the $\chi^2$ test statistic is not significant at
a 5\% level: in this example, its value is 6.221, and the corresponding
$p$ value is approximately $0.101$. Apparently, the relative size of the
tiles is insufficient to assess the ``strength'' of association
between two variables; again, information on the absolute size of the
residuals would be useful.
Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
that \emph{no} residual is bigger than 2 (or smaller than $-2$ for
negative ones, respectively) 
because simply no tile is shaded. In fact, as we can see from
the legend, the overall range of the residuals is smaller than $[-1.5,1.5]$.

@@

To visualize \textit{conditional} independence, we could use conditioning
plots: like in trellis displays, we produce a separate plot
for each level of a conditioning variable (in the case of two
or more conditioning variables, for each level combination). 
This can be illustrated using the famous admissions data of the University of
California at Berkley \citep{vcd:Bickel+Hammel+O'connell:1975} 
which is available in base \textsf{R}.
In this data, the question whether there is gender
discrimination at the university leads to the result that although women seem to
be disadvantaged at the aggregated level there is no gender discrimination
in the department strata---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
association plot in Figure \ref{fig:UCB-assoc-signif}.

@@

\section*{Addresses}

David Meyer\\
Achim Zeileis\\
E-mail: \texttt{\{David.Meyer,Achim.Zeileis\}@ci.tuwien.ac.at}\\
Institut f\"ur Statistik und Wahrscheinlichkeitstheorie\\
Technische Universit\"at Wien\\
Wiedner Hauptstra{\ss}e 8--10/1071\\
1040 Wien\\
\\
Kurt Hornik\\
E-mail: \texttt{Kurt.Hornik@wu-wien.ac.at}\\
Institut f\"ur Statistik\\
Wirtschaftsuniversit\"at Wien\\
Augasse 2--6\\
1090 Wien\\
